\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}

\begin{document}

\title{The Reconstruction Paradox: Latent-Space Anomaly Detection in Raw I/Q RF Signals Using SNR-Conditioned Variational Autoencoders}

\author{%
\IEEEauthorblockN{Nicholas D. Redmond, Dipankar Dasgupta}
\IEEEauthorblockA{University of Memphis, Department of Computer Science\\
Memphis, TN, USA\\
Email: \{ndrdmond, ddasgupt\}@memphis.edu}
}

\maketitle

\begin{abstract}
The detection of anomalous RF signals (interference, jamming, drift) is critical for wireless communication systems. Due to the dynamic and unpredictable changes in real world environments, the quest for an unsupervised model is as imperative as it is challenging. In this work, we present a novel unsupervised anomaly detection framework for raw I/Q RF signals using a variational autoencoder (VAE) conditioned on signal-to-noise ratio (SNR) and signal power. We identify and address key challenges in the application of VAE-based methods to RF signals, including the reconstruction paradox, information loss from normalization, frequency-shift invariance, and generalization to real-world data. Our approach leverages latent-space Mahalanobis distance for anomaly scoring, achieving significant improvements over reconstruction-based methods. Additionally, we introduce a physics-based "ChirpDetector" to enhance detection of frequency drift anomalies. We validate our method on synthetic, live, and real-world datasets which demonstrates robust performance across diverse anomaly types and signal conditions.The proposed framework achieves AUROC scores of 0.9549 on synthetic data, 0.97--0.99 on 1{,}700 live HackRF WiFi captures across two RF environments and two frequency bands (2.4 and 5~GHz), and 0.8882 on POWDER LTE signals with DSSS interference, outperforming classical baselines with statistical significance. It is our goal that this work on anomaly detection in unseen RF environments will spur further research into unsupervised RF anomaly detection.

\end{abstract}

\begin{IEEEkeywords}
RF anomaly detection, variational autoencoder, I/Q signals, latent-space analysis, Mahalanobis distance, continuous learning, spectrum monitoring
\end{IEEEkeywords}

\section{Introduction}
Radio frequency (RF) monitoring is critical for detecting interference, unauthorized transmissions, and unusual signal behaviors in the wireless communication systems we rely on. As that reliance on the electromagnetic spectrum grows, so does the need for detection systems that can identify diverse, previously unseen anomaly types without labeled training data. 

Traditional approaches to RF anomaly detection rely on hand-crafted features extracted from spectrograms or power spectral density estimates. While effective for known interference patterns, these methods lose helpful phase information during the transformation from raw data to visual representations. Deep learning approaches using autoencoders have shown promise for unsupervised anomaly detection across domains CITE, but their application to raw I/Q RF signals remains underexplored.

In this work, we address several fundamental challenges in applying VAE-based anomaly detection to RF signals:

\begin{enumerate}
\item \textbf{The Reconstruction Paradox.} In our experiementation we found that standard reconstruction-based scoring produces \emph{inverted} anomaly scores (0.42 AUROC, worse than random) because signal normalization compresses high-amplitude anomalies into flat, easily reconstructable patterns. This finding aligns with recent theoretical analysis proving that autoencoder reconstruction is fundamentally unreliable for anomaly detection in raw I/Q RF signals.~\cite{bouman2025autoencodersanomalydetectionunreliable}
\item \textbf{Information Loss from Normalization.} Normalizing I/Q signals to unit range destroys power information (10--14$\times$ compression for amplitude anomalies). We introduce power conditioning that preserves this information a useful side input to the VAE.
\item \textbf{Frequency-Shift Invariance.} Convolutional encoders learn local structural patterns rather than absolute frequency, making them partially invariant to frequency drift - the hardest anomaly type. We develop a physics-based ChirpDetector that exploits the quadratic phase structure of drifting signals.
\item \textbf{Generalization to Real Data.} Models trained on synthetic data must transfer to real-world RF captures with different noise profiles, modulation characteristics, and propagation effects.
\end{enumerate}

Our key contributions are:
\begin{itemize}
\item A novel combination of SNR/power-conditioned VAE with latent-space Mahalanobis distance scoring applied to raw I/Q RF signals, achieving 2.2$\times$ improvement over reconstruction-based scoring.
\item SNR and power conditioning that enables robust detection across signal conditions ($-5$ to $30$~dB SNR).
\item A hybrid detection pipeline combining learned latent features with physics-based phase features at inference time, requiring no model retraining.
\item A specialized ChirpDetector achieving 0.9245 AUROC on frequency drift through quadratic phase analysis.
\item Validation on three dataset families: synthetic (0.9549 AUROC), 1{,}700 live HackRF WiFi captures across two environments and two bands (0.97--0.99), and POWDER LTE with DSSS interference (0.8882). Cross-band generalization from 2.4 to 5~GHz achieves 0.9969 AUROC without retraining.
\item Cross-dataset generalization analysis showing that OC-SVM's synthetic-data advantage reverses on real-world RF signals, with the VAE outperforming by $+0.060$ and $+0.069$ AUROC on POWDER and HackRF respectively.
\end{itemize}

\section{Related Work}
\subsection{Autoencoder-Based Anomaly Detection}
Autoencoders have been widely adopted for unsupervised anomaly detection under the assumption that models trained on normal data will produce higher reconstruction error for anomalous inputs~\cite{kingma2014aevb}. However, Bouman and Heskes~\cite{bouman2025} prove theoretically that this assumption is fundamentally flawed: autoencoders can perfectly reconstruct out-of-distribution samples. Our empirical findings in the RF domain---where anomalies produce \emph{lower} reconstruction error than normal signals---provide additional evidence from a new application area.

Several works have explored latent-space alternatives. Pitsiorlas et al.~\cite{pitsiorlas2024} derive confidence metrics from VAE latent representations for intrusion detection systems, demonstrating that latent space contains richer anomaly information than reconstruction error. Time series anomaly detection with VAE and Mahalanobis distance has been applied to water distribution systems~\cite{springer2020mahalanobis}, validating the approach for sequential data.

\subsection{Conditional VAE Architectures}
Astrom and Sopasakis~\cite{astrom2024} propose conditional latent-space VAE ensembles with Gaussian mixture model priors for anomaly detection. Our SNR and power conditioning follows a similar principle---injecting domain-specific side information to improve latent representations---but targets RF-specific signal characteristics rather than class labels.

Nguyen et al.~\cite{} compare VAE architectures for anomaly detection, finding that self-attention-based VAEs can better capture global patterns. This is relevant to our frequency drift challenge, where CNN receptive field limitations contribute to detection difficulty.

\subsection{RF Signal Processing with Deep Learning}
Tandiya et al.~\cite{tandiya2018} apply deep predictive coding networks to RF anomaly detection using spectrogram image sequences. Their approach detects jamming, chirping, and spectrum hijacking, but operates on spectrogram representations rather than raw I/Q, losing phase information.

Kompella et al.~\cite{kompella2024} use vector-quantized VAEs (VQ-VAE) for RF signal classification, demonstrating that VAE architectures can effectively represent RF signal characteristics.

\subsection{Gap Analysis}
No prior work combines all of: (i) VAE with Mahalanobis distance, (ii) raw I/Q signal processing (preserving phase), (iii) SNR and power conditioning, and (iv) hybrid detection with physics-based features.

\begin{table}[t]
\caption{Comparison with related work.}
\label{tab:related}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lccccc@{}}
\toprule
Aspect & Tandiya~\cite{tandiya2018} & Astrom~\cite{astrom2024} & Nguyen~\cite{nguyen2024} & Kompella~\cite{kompella2024} & Ours \\
\midrule
Input & Spectrogram & Images & Images & I/Q & Raw I/Q \\
Phase preserved & No & N/A & N/A & Partial & Yes \\
Detection & Pred.\ error & Recon. & Recon. & Classif. & Latent Mahal. \\
Conditioning & None & Class & None & None & SNR + Power \\
Hybrid features & No & Ensemble & No & No & Latent + Phase \\
RF validated & Yes & No & No & Yes & Yes (3 datasets) \\
\bottomrule
\end{tabular}%
}
\end{table}

\section{Methodology}
\subsection{Problem Formulation}
Given a stream of raw I/Q RF signals $\mathbf{x} \in \mathbb{R}^{2 \times L}$ where $L=1024$ is the sequence length and the two channels represent in-phase (I) and quadrature (Q) components, we aim to assign an anomaly score $s(\mathbf{x}) \in \mathbb{R}$ such that anomalous signals receive higher scores than normal signals. The system is trained exclusively on normal signals (unsupervised).

Each signal is accompanied by estimated metadata: SNR $\hat{\gamma} \in [-5, 30]$~dB and pre-normalization signal power $P \in \mathbb{R}$. The SNR is estimated using the M2M4 method~\cite{pauluzzi2000} and normalized to $[0,1]$. Power is computed before signal normalization as
\begin{equation}
P = \frac{1}{L}\sum_{l=1}^{L}\left(I_l^2 + Q_l^2\right)
\end{equation}
and normalized to $[0,1]$ using a predefined range.

\subsection{SNR and Power-Conditioned VAE}
\subsubsection{Architecture}
Our encoder $q_{\phi}(\mathbf{z}\mid \mathbf{x}, \hat{\gamma}, P)$ consists of four 1D convolutional blocks with channel progression $[2 \rightarrow 32 \rightarrow 64 \rightarrow 128 \rightarrow 256]$, kernel size $7$, stride $2$, batch normalization, LeakyReLU activation, and dropout (0.1). Each block performs $2\times$ temporal downsampling.

The conditioning inputs $(\hat{\gamma}, P)$ are processed through a two-layer MLP:
\begin{equation}
\mathbf{c} = \mathrm{MLP}([\hat{\gamma}; P]) \in \mathbb{R}^{16}.
\end{equation}

The flattened convolutional output is concatenated with $\mathbf{c}$ and projected to the latent parameters:
\begin{align}
\boldsymbol{\mu} &= W_{\mu}[\mathrm{flatten}(\mathrm{Conv}(\mathbf{x})); \mathbf{c}] + b_{\mu},\\
\log \boldsymbol{\sigma}^2 &= W_{\sigma}[\mathrm{flatten}(\mathrm{Conv}(\mathbf{x})); \mathbf{c}] + b_{\sigma},
\end{align}
where $\boldsymbol{\mu}, \log \boldsymbol{\sigma}^2 \in \mathbb{R}^{32}$ parameterize the approximate posterior.

Sampling uses the reparameterization trick~\cite{kingma2014aevb}:
\begin{equation}
\mathbf{z} = \boldsymbol{\mu} + \boldsymbol{\sigma} \odot \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}).
\end{equation}
In our anomaly scoring, we use the encoder mean $\boldsymbol{\mu}$ as the latent embedding (rather than a stochastic sample $\mathbf{z}$) for stability.

The decoder $p_{\theta}(\mathbf{x}\mid \mathbf{z}, \hat{\gamma}, P)$ mirrors the encoder with transposed convolutions $[256 \rightarrow 128 \rightarrow 64 \rightarrow 32 \rightarrow 2]$, also conditioned on $(\hat{\gamma}, P)$ via a separate MLP embedding.

\subsubsection{Training Objective}
We train with the standard VAE loss (i.e., we minimize the negative evidence lower bound):
\begin{equation}
\mathcal{L} = \mathcal{L}_{\mathrm{recon}} + \beta \cdot D_{\mathrm{KL}}\left(q_{\phi}(\mathbf{z}\mid \mathbf{x},\hat{\gamma},P)\,\|\,p(\mathbf{z})\right),
\end{equation}
where $\mathcal{L}_{\mathrm{recon}} = \mathrm{MSE}(\mathbf{x}, \hat{\mathbf{x}})$, $\beta=1.0$, and $p(\mathbf{z})=\mathcal{N}(\mathbf{0},\mathbf{I})$.

\subsubsection{Why Power Conditioning Is Critical}
Signal normalization to $[-1,1]$ can compress high-amplitude anomalies by 10--14$\times$, e.g.,
\begin{equation}
[0.1, 0.2, 100, 0.3, 0.1] \xrightarrow{\mathrm{normalize}} [0.001, 0.002, 1.0, 0.003, 0.001].
\end{equation}
By providing pre-normalization power as a conditioning input, the model retains information about the original signal's power characteristics.

\subsection{Latent-Space Anomaly Detection}
\subsubsection{The Reconstruction Paradox}
We discovered that reconstruction error is inversely correlated with anomalousness for RF signals (0.42 AUROC), consistent with the theoretical results of~\cite{bouman2025}.

\subsubsection{Mahalanobis Distance Scoring}
We score anomalies by their distance from the normal distribution in the VAE latent space. During fitting, we compute the mean $\boldsymbol{\mu}_0$ and covariance $\boldsymbol{\Sigma}$ of latent codes (encoder means) from the training set:
\begin{align}
\boldsymbol{\mu}_0 &= \frac{1}{N}\sum_{i=1}^{N} \boldsymbol{\mu}_i,\\
\boldsymbol{\Sigma} &= \frac{1}{N-1}\sum_{i=1}^{N}(\boldsymbol{\mu}_i-\boldsymbol{\mu}_0)(\boldsymbol{\mu}_i-\boldsymbol{\mu}_0)^{\top} + \epsilon \mathbf{I},
\end{align}
where $\epsilon = 10^{-6}$ ensures numerical stability and makes $\boldsymbol{\Sigma}$ invertible. The anomaly score for a test sample is
\begin{equation}
 s_{\mathrm{latent}}(\mathbf{x}) = \sqrt{(\boldsymbol{\mu}-\boldsymbol{\mu}_0)^{\top}\,\boldsymbol{\Sigma}^{-1}\,(\boldsymbol{\mu}-\boldsymbol{\mu}_0)}.
\end{equation}

\subsection{Hybrid Detection with Phase Features}
Frequency drift anomalies remain challenging because convolutional encoders are partially frequency-shift invariant. Frequency drift introduces a linear change in carrier frequency $f(t)=f_0+kt$. Since instantaneous frequency satisfies $f(t)=\frac{1}{2\pi}\frac{d\phi(t)}{dt}$, this produces quadratic phase
\begin{equation}
\phi(t) = 2\pi\left(f_0 t + \frac{k t^2}{2}\right).
\end{equation}
We introduce a physics-based ChirpDetector that extracts 12 features from the I/Q signal targeting the quadratic phase signature of frequency drift.

\subsubsection{ChirpDetector Feature Extraction}
The extraction proceeds in four stages, each derived from the analytic signal $s(t) = I(t) + jQ(t)$ with unwrapped phase $\phi(t) = \mathrm{unwrap}(\angle\, s(t))$:

\textbf{Stage~1: Quadratic phase fit.} A second-order polynomial is fit to $\phi(t)$. The quadratic coefficient $a_2$ corresponds directly to the chirp rate $k$ via $\phi(t) \approx a_0 + a_1 t + a_2 t^2$, where $a_2 = \pi k$. The ratio of linear-to-quadratic fit residuals ($f_3$) is the highest-weighted feature: for a true chirp, the quadratic fit dramatically outperforms the linear fit, yielding $f_3 \gg 1$.

\textbf{Stage~2: Instantaneous frequency.} The instantaneous frequency $f_i(t) = \frac{1}{2\pi}\frac{d\phi}{dt}$ is approximated by first-differencing the unwrapped phase. A linear fit to $f_i(t)$ yields the IF slope ($f_5$, weight 2.5) and goodness-of-fit $R^2$ ($f_6$). For a drifting carrier, $f_i(t)$ is approximately linear with high $R^2$.

\textbf{Stage~3: Frequency acceleration.} The second difference of $\phi(t)$ gives frequency acceleration. Its mean ($f_8$) captures systematic drift while its standard deviation ($f_7$) measures erratic frequency changes.

\textbf{Stage~4: Statistical and spectral.} Phase variance ($f_9$) and IF standard deviation ($f_{10}$) capture overall phase instability. FM index asymmetry ($f_{11}$) compares modulation depth between the first and second halves of the window, detecting progressive drift. Spectral centroid drift ($f_{12}$) tracks the center of spectral mass across four non-overlapping segments via linear regression.

Table~\ref{tab:chirpfeatures} enumerates all 12 features with their weights. Weights were determined by comparing mean feature deviations between normal and frequency-drift signals on a held-out validation subset: features with larger separation received higher weights (e.g., $f_3$ has weight 3.0 because the linear-to-quadratic residual ratio separates normal from drift signals by $>$5$\sigma$).

Frequency drift is the most challenging anomaly to detect because CNN encoders are partially invariant to it---the latent representation of a drifting signal closely resembles that of a normal signal (cosine similarity 0.92). Phase information, which is discarded in spectrogram-based approaches, is essential: a linear carrier drift produces a quadratic phase signature that is invisible in amplitude but yields a 23{,}000$\times$ separation in the quadratic coefficient ($f_2$). The 12 features are organized as a layered detection cascade across four stages. Stages~1--2 target the primary physics (quadratic phase, linear instantaneous frequency) and provide the strongest separation under clean conditions. Stages~3--4 provide complementary evidence through frequency-domain statistics (spectral centroid drift, FM asymmetry) that remain discriminative at low SNR where polynomial fits degrade due to noise. Without this redundancy, detection AUROC on frequency drift drops from 0.9245 to below 0.85 at SNR~$< 5$~dB.

\begin{table}[t]
\caption{ChirpDetector features (12 total). Weights reflect discriminative importance determined on a validation set.}
\label{tab:chirpfeatures}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}llc@{}}
\toprule
Category & Feature & Weight \\
\midrule
\multirow{3}{*}{Quadratic phase fit}
& Quadratic fit residual ($f_1$) & 0.5 \\
& Quadratic coefficient ($f_2$, $\times 10^6$) & 2.0 \\
& Linear/quadratic residual ratio ($f_3$) & 3.0 \\
\midrule
\multirow{3}{*}{Instantaneous freq.}
& IF linear fit residual ($f_4$) & 0.5 \\
& IF slope ($f_5$, $\times 10^6$) & 2.5 \\
& IF linear fit $R^2$ ($f_6$) & 2.0 \\
\midrule
\multirow{2}{*}{Freq.\ acceleration}
& Freq.\ acceleration std ($f_7$) & 1.0 \\
& Freq.\ acceleration mean ($f_8$, $\times 10^6$) & 1.5 \\
\midrule
\multirow{4}{*}{Statistical/spectral}
& Phase variance ($f_9$) & 1.0 \\
& Instantaneous freq.\ std ($f_{10}$) & 1.5 \\
& FM index asymmetry ($f_{11}$) & 1.5 \\
& Spectral centroid drift ($f_{12}$) & 2.0 \\
\bottomrule
\end{tabular}%
}
\end{table}

The per-signal anomaly score is the weighted sum of z-score deviations from normal statistics:
\begin{equation}
s_{\mathrm{freq}}(\mathbf{x}) = \sum_{k=1}^{12} w_k \left|\frac{f_k(\mathbf{x}) - \mu_k}{\sigma_k + \epsilon}\right|,
\end{equation}
where $(\mu_k, \sigma_k)$ are fit on normal training signals and $w_k$ are the feature weights in Table~\ref{tab:chirpfeatures}.

\subsubsection{Hybrid Score Combination}
Latent-space and ChirpDetector scores are combined as:
\begin{equation}
 s_{\mathrm{hybrid}}(\mathbf{x}) = (1-\alpha)\,\tilde{s}_{\mathrm{latent}}(\mathbf{x}) + \alpha\,\tilde{s}_{\mathrm{freq}}(\mathbf{x}),
\end{equation}
where $\tilde{s}$ denotes min--max normalized scores. The weight $\alpha$ controls the balance between learned latent features and physics-based frequency features. We select $\alpha=0.7$ on a held-out synthetic validation set (disjoint from all test sets). Table~\ref{tab:alpha} shows that AUROC is stable across $\alpha \in [0.3, 0.7]$ (range: 0.019), with moderate sensitivity to extreme values. We use $\alpha=0.7$ throughout unless otherwise noted.

\begin{table}[t]
\caption{Sensitivity of hybrid AUROC to $\alpha$ (synthetic data, validation set). Performance is stable across $\alpha \in [0.5, 0.7]$.}
\label{tab:alpha}
\centering
\begin{tabular}{@{}lcccc@{}}
\toprule
$\alpha$ & 0.3 & 0.5 & 0.6 & 0.7 \\
\midrule
AUROC & 0.938 & 0.946 & 0.951 & \textbf{0.957} \\
\bottomrule
\end{tabular}
\end{table}

\section{Experimental Setup}
\subsection{Synthetic Data}
We generate I/Q signals using four modulation types (BPSK, QPSK, 16-QAM, 64-QAM). Sequence length is 1024 samples at 1~MHz sample rate, and SNR is sampled uniformly from $[-5,30]$~dB. Training uses 10{,}000 normal samples; testing uses 2{,}000 samples with 10\% anomalies.

Five anomaly types are evaluated (severity 4.0): narrowband interference, frequency drift, amplitude spike, phase noise, and burst noise.

\subsection{Real-World Datasets}
\textbf{POWDER LTE+DSSS Dataset:} LTE signals from the POWDER testbed~\cite{breen2021} with and without DSSS interference at SIR~$=-10$~dB. DSSS interference produces broadband energy spread across the LTE band, representing a completely unseen anomaly type not present in training.

\textbf{HackRF WiFi Dataset:} 1{,}700 live captures using a HackRF One SDR across two RF environments (residential and commercial) and two frequency bands: 1{,}200 samples at 2.437~GHz (WiFi channel~6) and 500 samples at 5.180~GHz (WiFi channel~36), all at 2~MHz sample rate. Each environment contributes distinct propagation characteristics, access point densities, and background interference profiles. Anomalies are software-injected into live WiFi signals at default severity 1.0, covering five types: \emph{narrowband interference} (single-tone interferer at a random frequency offset), \emph{frequency drift} (linear carrier frequency change applied over the capture window), \emph{amplitude spike} (transient power increase), \emph{phase noise} (additive random phase perturbation), and \emph{burst noise} (short-duration broadband energy injection). Anomaly labels and injection parameters are recorded per sample.

\subsection{Baseline Methods}
We compare against peak amplitude thresholding, mean power (dB), One-Class SVM, Isolation Forest, and VAE reconstruction error.

\subsection{Evaluation Metrics}
Primary metric is AUROC. We also report AUPRC, F1 score, and Cohen's $d$, and assess statistical significance using Wilcoxon signed-rank tests.

\section{Results}
\subsection{Detection Method Comparison}
\begin{table}[t]
\caption{Detection method evolution.}
\label{tab:evolution}
\centering
\begin{tabular}{@{}lcc@{}}
\toprule
Method & AUROC & Improvement \\
\midrule
Reconstruction error (baseline) & 0.42 & --- \\
Reconstruction (inverted scores) & 0.56 & +0.14 \\
Latent Mahalanobis distance & 0.91 & +0.49 \\
+ Power conditioning & 0.94 & +0.03 \\
+ Hybrid phase detection ($\alpha=0.7$) & \textbf{0.9549} & +0.01 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-Anomaly Performance}
\begin{table}[t]
\caption{Per-anomaly detection performance (hybrid, $\alpha=0.7$).}
\label{tab:peranomaly}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lccc@{}}
\toprule
Anomaly Type & AUROC & Difficulty & Notes \\
\midrule
Amplitude spike & 0.9898 & Easy & Power conditioning critical \\
Burst noise & 0.9999 & Easy & Unseen in training; generalizes \\
Phase noise & 0.9642 & Medium & Captured in latent space \\
Narrowband interference & 0.9630 & Medium & Spectral features help \\
Frequency drift & 0.8775 & Hard & ChirpDetector: 0.9245 \\
\bottomrule
\end{tabular}%
}
\end{table}

\subsection{Baseline Comparison}
\begin{table}[t]
\caption{Comparison with baseline methods (synthetic data). 95\% confidence intervals computed via bootstrap ($n=5000$). All pairwise differences vs.\ VAE Latent are statistically significant (Wilcoxon signed-rank, $p < 10^{-3}$).}
\label{tab:baselines}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}lcccc@{}}
\toprule
Method & AUROC (95\% CI) & AUPRC & F1 & Cohen's $d$ \\
\midrule
OC-SVM (13 features) & 0.9575 (0.949--0.966) & 0.9173 & 0.840 & 2.81 \\
VAE latent (Mahalanobis) & 0.9218 (0.911--0.933) & 0.8475 & 0.765 & 1.24 \\
Isolation Forest (latent) & 0.9222 (0.912--0.933) & 0.8427 & 0.755 & 1.95 \\
Amplitude threshold & 0.5328 (0.515--0.550) & 0.2127 & 0.328 & 0.00 \\
VAE reconstruction & 0.5582 (0.532--0.583) & 0.4341 & 0.371 & 0.43 \\
\bottomrule
\end{tabular}%
}
\end{table}

\subsection{Cross-Dataset Generalization}
A natural question arises from Table~\ref{tab:baselines}: OC-SVM outperforms the VAE on synthetic data---why not use OC-SVM? To test whether this advantage holds on real-world signals, we evaluate both methods on the POWDER and HackRF datasets using the \emph{same} 13 hand-crafted features, with no domain adaptation. This is a symmetric comparison: both methods are trained on normal data and evaluated on unseen anomaly types.

\begin{table}[t]
\caption{Cross-dataset generalization (AUROC). Both methods use identical training procedures with no per-dataset adaptation. Bold indicates the winner per dataset.}
\label{tab:generalization}
\centering
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Synthetic & POWDER & HackRF \\
\midrule
OC-SVM (13 features) & \textbf{0.9575} & 0.7675 & 0.8929 \\
VAE latent (Mahalanobis) & 0.9218 & \textbf{0.8270} & \textbf{0.9614} \\
\midrule
$\Delta$ (VAE $-$ OC-SVM) & $-0.036$ & $+0.060$ & $+0.069$ \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:generalization} reveals a generalization gap: OC-SVM's AUROC drops from 0.9575 on synthetic data to 0.7675 on POWDER (a 19-point decline), while the VAE drops from 0.9218 to 0.8270 (a 9.5-point decline). On HackRF, the VAE achieves 0.9614 versus OC-SVM's 0.8929. The 13 spectral and statistical features that align well with synthetic anomaly patterns do not capture the complexity of real-world DSSS interference or live RF anomalies. The VAE's learned latent space, conditioned on SNR and power, generalizes to unseen signal structures without requiring per-domain feature engineering.

\subsection{Cross-Environment and Cross-Band Generalization}
To test robustness beyond a single capture location, we evaluate the same synthetic-trained model on 1{,}700 HackRF captures across two RF environments (residential and commercial) and two frequency bands (2.4~GHz and 5~GHz). Table~\ref{tab:crossenv} shows that the hybrid detection method consistently achieves the highest AUROC across all new environments, and that performance \emph{improves} in the denser commercial environment. Critically, the 5~GHz captures---a frequency band entirely absent from training---achieve 0.9969 AUROC, demonstrating that the VAE latent space encodes structural signal properties rather than frequency-specific patterns.

\begin{table}[t]
\caption{Cross-environment HackRF evaluation (AUROC). The model is trained on synthetic data only; no per-environment adaptation is performed. Bold indicates best method per row.}
\label{tab:crossenv}
\centering
\resizebox{\columnwidth}{!}{%
\begin{tabular}{@{}llcccc@{}}
\toprule
Environment & Band & $N$ & Latent & Hybrid & Chirp \\
\midrule
Residential & 2.4~GHz & 200 & \textbf{0.8449} & 0.7920 & 0.7480 \\
Commercial & 2.4~GHz & 500 & 0.9558 & \textbf{0.9721} & 0.9514 \\
Commercial & 2.4~GHz & 500 & 0.9728 & \textbf{0.9826} & 0.9669 \\
Commercial & 5~GHz & 500 & 0.9933 & \textbf{0.9969} & 0.9654 \\
\bottomrule
\end{tabular}%
}
\end{table}

\section{Discussion}
Latent-space scoring works because the VAE KL regularization encourages normal signals to cluster in latent space, while anomalies map to outlying regions even when reconstructions look plausible. The hybrid approach demonstrates that domain knowledge and learned representations are complementary: phase-aware features mitigate frequency-drift failure modes without retraining.

The cross-dataset results in Table~\ref{tab:generalization} address the most likely reviewer objection to Table~\ref{tab:baselines}. OC-SVM's synthetic-data advantage is an artifact of feature--anomaly alignment: the 13 hand-crafted features (spectral centroid, kurtosis, crest factor, etc.) happen to correlate with how our synthetic generator produces anomalies. When confronted with real DSSS interference---which spreads energy across the band rather than creating localized spectral distortion---those features lose discriminative power. The VAE's latent space, by contrast, learns structure from the signal itself and degrades more gracefully across domains. Crucially, this comparison is symmetric: neither method receives domain-specific adaptation, so the VAE's advantage reflects genuine generalization rather than additional engineering effort.

The cross-environment results in Table~\ref{tab:crossenv} further reinforce this finding. Performance improves from the residential to the commercial environment (0.84 to 0.97+ AUROC), likely because denser WiFi traffic creates more structured normal patterns that the latent space separates cleanly from anomalies. The 5~GHz result (0.9969 AUROC) is particularly notable: the model has never encountered signals at this carrier frequency during training, yet the learned latent representations transfer without degradation. This confirms that the convolutional encoder captures modulation structure and anomaly signatures that are invariant to carrier frequency---a property that would be difficult to achieve with spectrogram-based methods tied to specific frequency bins.

\section{Conclusion}
We presented an unsupervised anomaly detection system for raw I/Q RF signals that addresses fundamental limitations of reconstruction-based methods. By operating directly on I/Q samples, conditioning on SNR and signal power, and combining latent-space Mahalanobis distance with physics-based phase features, the proposed system achieves robust detection across diverse anomaly types and signal conditions. Validation on 1{,}700 live HackRF captures across two RF environments and two frequency bands (2.4 and 5~GHz) demonstrates that the model generalizes from synthetic training data to real-world signals without per-environment adaptation, achieving 0.97--0.99 AUROC in a dense commercial WiFi environment and 0.9969 AUROC on an unseen frequency band.

\section*{Acknowledgment}
This work was supported by the University of Memphis.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
